|

### 2025-07-14 output/IMPS-2025-slideshow.qmd
Add new "oblique axes" slides

>>>>

## Oblique axes {.smaller}

:::::: columns
::: {.column width="33%"}
```{r}
#| fig-width:  5.5
#| fig-height: 5.5
thetas_transf |>
  ggplot(mapping = aes(trait_1, trait_2)) +
  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
  scale_x_continuous(
    minor_breaks = NULL,
    labels = \(breaks) '' |> rep(length(breaks)),
    limits = c(-4, 4),
    expand = expansion(),
    name   = NULL,
    sec.axis = dup_axis(name = "{*&theta;*~1~, *&theta;*~2~}", labels = NULL)
  ) +
  scale_y_continuous(
    minor_breaks = NULL,
    labels = \(breaks) '' |> rep(length(breaks)),
    limits = c(-4, 4),
    expand = expansion(),
    name   = NULL
  ) +
  coord_fixed() +
  theme(
    axis.ticks   = element_blank(),
    axis.line    = element_blank(),
    axis.title   = element_markdown(),
    axis.title.y = element_markdown(),
    panel.grid   = element_blank()
  )
```
:::

::: {.column .fragment width="33%" fragment-index="1"}
```{r question-mark}
fa(
  "question",
  fill           = "#929acb",
  height         = "200px",
  vertical_align = "-10em",
  margin_left    = "4em"
)
```
:::

::: {.column .fragment width="33%" fragment-index="2"}
```{r corr-traits-transf-example}
#| fig-width:  5.5
#| fig-height: 5.5
```
:::
::::::

<center>

::: {.fragment fragment-index="2"}
Which coordinate system *θ* makes *θ\** "spherical"?
:::

</center>

::: notes
So, if the basis matrix of $\mathbf{\theta^*}$, which has an identity covariance matrix, is an identity covariance matrix, what is $\mathbf{P}$, the basis matrix of $\mathbf{\theta}$.

In other words,...

...what is the coordinate system of $\mathbf{\theta}$...

...that makes $\mathbf{\theta^*}$ spherical?
:::

## Oblique axes

```{r create-grid-function}
create_grid <- function(basis,
                        x_limits = c(-4, 4),
                        y_limits = c(-4, 4),
                        break_step = 2) {
  # Argument checking and formatting: ----

  ## Basis vectors tangents and norms
  vec_tan <- basis[2, ] / basis[1, ]

  # Main: ----

  # Define grid limits as a function of the axes limits (in canonical metric):
  grid_box <- expand_grid(x = x_limits, y = y_limits) |> mutate(
    x_lim = if_else(x == x_limits[1], "inf", "sup"),
    y_lim = if_else(y == y_limits[1], "inf", "sup")
  )

  # Select delimiting grid vertices, based on the basis vector directions
  grid_bounds <- bind_rows(
    h = grid_box |> mutate(slope = vec_tan[1], limit = y_lim),
    v = grid_box |> mutate(slope = vec_tan[2], limit = x_lim),
    .id = "grid"
  ) |>
    filter(xor(slope > 0, x_lim == y_lim)) |>
    mutate(limit_intercept = y - slope * x) # Compute their intercept

  # Transformed (1, 1), expressed in the canonical basis
  unit_point <- basis %*% c(1, 1) |> drop() |>
    as_tibble(.name_repair = "minimal") |>
    add_column(name = c('x', 'y') |> paste0('_unit')) |>
    pivot_wider(values_from = value)

  # Grid specifications, in canonical basis
  grid_specs <- grid_bounds |>
    bind_cols(unit_point) |>
    mutate(unit_intercept = y_unit - slope * x_unit) |>
    select(-ends_with("unit")) |>
    mutate(
      n_units  = trunc(limit_intercept / unit_intercept),
      n_breaks = n_units %/% break_step * break_step
    )

  # Create vertical grid data, if the second basis vector is completely vertical
  if (is.infinite(vec_tan[2])) {

    # Create vertical grid data
    v_grid <- grid_bounds |>
      filter(grid == "v") |>
      mutate(
        unit     = basis[1, 1],
        n_units  = trunc(x / unit),
        n_breaks = n_units %/% break_step * break_step
      ) |>
      select(grid, limit, n_breaks, unit) |>
      pivot_wider(names_from = "limit", values_from = n_breaks) |>
      reframe(coord = seq(from = inf, to = sup, by = break_step) * unit) |>
      mutate(
        grid      = 'v',
        slope     = Inf,
        intercept = NA_real_,
        label_pos = coord,
        ref       = as.character(!coord)
      )

    # Filter out non-valid grid values
    grid_specs <- grid_specs |> filter(!is.infinite(slope))
  }

  # Create grid data
  grid <- grid_specs |>
    select(grid, limit, slope, unit_intercept, n_breaks) |>
    pivot_wider(names_from = limit, values_from = n_breaks) |>
    group_by(grid) |>
    reframe(
      slope,
      coord = seq(from = inf, to = sup, by = break_step),
      intercept = coord * unit_intercept
    ) |>
    mutate(
      label_pos = if_else(
        grid == "h",
         intercept   + x_limits[1] * vec_tan[1],
        (y_limits[1] - intercept)  / vec_tan[2]
      ),
      ref = as.character(!intercept)
    )

    if (is.infinite(vec_tan[2])) return(grid |> bind_rows(v_grid))

    grid
}

plot_grid <- function(grid,
                      x_limits = c(-4, 4),
                      y_limits = c(-4, 4),
                      break_step = 2,
                      linetype_axis = 'solid',
                      linetype_grid = 'dashed') {

  # Argument checking and formatting: ----

  ## Grid plot linetypes
  linetypes <- c(`TRUE` = linetype_axis, `FALSE` = linetype_grid)

  ## Basis vectors tangents and norms
  vec_tan <- grid |> distinct(grid, slope) |> deframe()

  ## Axis title positions
  y_axis_titpos <- (y_limits[2] / vec_tan[2] - x_limits[1]) /
    (x_limits[2] - x_limits[1])
  x_axis_titpos <- (x_limits[2] * vec_tan[1] - y_limits[1]) /
    (y_limits[2] - y_limits[1])

  # Main: ----

  # Assign x axis label values and coordinates (to account for the vertical
  #   special case)
  x_labels <- grid |> filter(grid == "v") |> select(coord, label_pos)

  # Create vertical grid geometry if the second vector is completely vertical
  if (is.infinite(vec_tan[2])) {

    geom_vgrid <- geom_vline(
      mapping = aes(xintercept = coord, linetype = ref),
      data = grid |> filter(grid == "v")
    )

    grid <- grid |> filter(grid == "h")
  }

  grid_plot <- grid |>
    ggplot() +
    geom_abline(
      mapping = aes(
        intercept = intercept,
        slope     = slope,
        linetype  = ref
      )
    )

  if (is.infinite(vec_tan[2])) grid_plot <- grid_plot + geom_vgrid

  grid_plot +
    scale_x_continuous(
      minor_breaks = NULL,
      breaks = x_labels |> pull(label_pos),
      labels = x_labels |> pull(coord),
      limits = x_limits,
      expand = expansion(),
      name   = NULL,
      sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
    ) +
    scale_y_continuous(
      minor_breaks = NULL,
      breaks = grid |> filter(grid == "h") |> pull(label_pos),
      labels = grid |> filter(grid == "h") |> pull(coord),
      limits = y_limits,
      expand = expansion(),
      name   = NULL,
      sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
    ) +
    scale_linetype_manual(values = linetypes, guide = NULL) +
    coord_fixed() +
    theme(
      axis.ticks   = element_blank(),
      axis.line    = element_blank(),
      axis.title.x       = element_markdown(hjust = y_axis_titpos),
      axis.title.y.right = element_markdown(vjust = x_axis_titpos)
    )
}

transform_grid <- function(basis,
                           x_limits = c(-4, 4),
                           y_limits = c(-4, 4),
                           break_step = 2,
                           linetype_axis = 'solid',
                           linetype_grid = 'dashed') {

  # Argument checking and formatting: ----

  # Main: ----
  basis |>
    create_grid(
      x_limits   = x_limits,
      y_limits   = y_limits,
      break_step = break_step
    ) |>
    plot_grid(
      x_limits   = x_limits,
      y_limits   = y_limits,
      break_step = break_step,
      linetype_axis = 'solid',
      linetype_grid = 'dashed'
    )
}
```

:::::: columns
::: {.column width="33%"}
<center>A</center>

```{r grid-orthogonal-rotation}
#| fig-width: 5
rot_basis <- cbind(
  c(sqrt(.75), -sqrt(.25)),
  c(sqrt(.25), sqrt(.75))
)

rot_grid <- transform_grid(rot_basis)

rot_grid
```
:::

::: {.column width="33%"}
<center>B</center>

```{r grid-ts-basis}
#| fig-width: 5
ts_basis <- cbind(
  c(1, 0),
  c(sqrt(.25), sqrt(.75))
)

ts_grid <- transform_grid(ts_basis)

ts_grid
```
:::

::: {.column width="33%"}
<center>C</center>

```{r grid-ls-basis}
#| fig-width: 5
ls_basis <- t(solve(ts_basis))

ls_grid <- transform_grid(ls_basis)

ls_grid
```
:::
::::::

::: fragment
[Answers in 3, 2, 1...](https://wall.sli.do/event/dPunB4BMx3S8hfSjaFSuFh?section=ae77dd00-34fb-4600-9677-337a65b04b83)
:::

::: notes
So, here are three possible options; one of them is the correct one.

Take a few seconds to answer and we will see next which is the correct one.
:::

## Oblique axes

:::::::::: columns
::: {.column width="33%"}
<center>A</center>

```{r scatter-orthogonal-rotation}
#| fig-width: 5
thetas_transf <- thetas |>
  group_by(i) |>
  pivot_longer(starts_with("trait")) |>
  mutate(value = rot_basis %*% value |> drop()) |>
  pivot_wider() |>
  ungroup()

rot_grid +
  geom_point(
    data    = thetas_transf,
    mapping = aes(trait_1, trait_2),
    alpha   = .33,
    shape   = 16,
    colour = I(PALETTE_BLUE["normal"])
  )
```
:::

::::: {.column width="33%"}
<center>B</center>

:::: r-stack
```{r grid-ts-basis}
#| fig-width: 5
```

::: fragment
```{r scatter-ts-basis}
#| fig-width: 5
thetas_transf <- thetas |>
  group_by(i) |>
  pivot_longer(starts_with("trait")) |>
  mutate(value = ts_basis %*% value |> drop()) |>
  pivot_wider() |>
  ungroup()

ts_grid +
  geom_point(
    data    = thetas_transf,
    mapping = aes(trait_1, trait_2),
    alpha   = .33,
    shape   = 16,
    colour = I(PALETTE_BLUE["normal"])
  )
```
:::
::::
:::::

::::: {.column width="33%"}
<center>C</center>

:::: r-stack
```{r grid-ls-basis}
#| fig-width: 5
```

::: fragment
```{r scatter-ls-basis}
#| fig-width: 5
thetas_transf <- thetas |>
  group_by(i) |>
  pivot_longer(starts_with("trait")) |>
  mutate(value = ls_basis %*% value |> drop()) |>
  pivot_wider() |>
  ungroup()

ls_scatter <- ls_grid +
  geom_point(
    data    = thetas_transf,
    mapping = aes(trait_1, trait_2),
    alpha   = .33,
    shape   = 16,
    colour = I(PALETTE_BLUE["normal"])
  )

ls_scatter
```
:::
::::
:::::
::::::::::

::: notes
A is pretty clearly not the answer, as it is just an orthogonal rotation.

Now, if you chose B, you probably thought that the axis units should not change (because the variances are unitary), and that the axes should form an angle which cosine is the correlation, that is, 0.5 (a 60º angle).

In that case, you think exactly like I do.

But if you picture this transformation in your mind, you will notice that when you make the axes closer, you will be squeezing the points in the distribution, making them look even more like a "sausage", so to speak.

The resulting distribution is this; it took us hundreds of hours of work, during more than two years to realise this.

So if you want to actually "stretch" the cloud of points in the distribution, you need to pull the axes apart,...

...as in the coordinate axes C.
:::
